### 2013-06-08 doc/pgas13/flat-combining.tex
Write abstract, add placeholder for conclusion.
edits to the sequential consistency

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.5\textwidth]{figs/fc_shared_mem.pdf}
  \caption{\emph{Flat combining in shared memory.}
    To access the shared stack, each thread adds its request to the publication list (1). Thread 4 acquires the stack's lock and becomes the combiner (2), while the remaining threads spin waiting for their request to be satisfied. The combiner walks the publication list from the head, matching up Thread 3's push with Thread 2's pop on its own private stack (3). The 2 remaining pushes are added to the top of the shared stack (4). Finally, the top is updated, and Thread 4 releases the lock and continues its normal execution.
  }
  \label{fig:fc_shared_mem}
\end{figure}

@@

\paragraph{Sequential Consistency}
The sequential consistency guarantee is that within a task, operations will be in task order, so to enforce ordering between operations in different tasks, one task must perform an operation that another one can observe. An example is shown in Figure~\ref{fig:sync}.\TODO{make sync example}
The Grappa memory model is in the style of the C++ memory model~\cite{boehm:drf0,N2480,N2800} memory model (guaranteeing sequential consistency for data-race free programs). Being conservative, delegate operations block the calling worker until they have completed globally, ensuring they can be used for synchronization. As such, delegate operations within a task are guaranteed to be globally visible in program order.
Because it is not immediately obvious that this distributed version of flat combining preserves sequential consistency, here we will make arguments for why it does.

@@

The correctness of combining set and map operations is more nuanced.
Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally.
Therefore, a batch of these operations can be committed globally in parallel.
If an insert finds that its key is already in the combiner, it does not need to send its own message. However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened.
In the same way, lookups can piggy-back on other lookups of the same key.
It is tempting to allow a lookup to be satisfied locally by inspecting the keys to be inserted.
However, this allows the local order in which keys were inserted to be observed.
If this is allowed, then that same order must be respected globally, forcing each batch to ensure atomicity from all other cores' batches. This is prohibitively expensive, so a cheaper solution is chosen: lookups must get their values from the global data structure, but can be performed in parallel in the same way as inserts.

@@

These requirements only guarantee sequential consistency for a data structure in isolation. Atomicity and data race freedom in the case of multiple data structures must be guaranteed externally, just as in any multithreaded shared-memory program.

>>>>

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{figs/fc_shared_mem.pdf}
  \caption{\emph{Flat combining in shared memory.}
    To access the shared stack, each thread adds its request to the publication list (1). Thread 4 acquires the stack's lock and becomes the combiner (2), while the remaining threads spin waiting for their request to be satisfied. The combiner walks the publication list from the head, matching up Thread 3's push with Thread 2's pop on its own private stack (3). The 2 remaining pushes are added to the top of the shared stack (4). Finally, the top is updated, and Thread 4 releases the lock and continues its normal execution.
  }
  \label{fig:fc_shared_mem}
\end{figure}

@@

\paragraph{Memory Consistency Model}

In the context of this work, sequential consistency guarantees that within a
task, operations will be in task order and that all tasks will observe the
same global order of operations. An example is shown in
Figure~\ref{fig:sync}.\TODO{make sync example}

The Grappa memory model is essentially the same as the C++ memory
model~\cite{boehm:drf0,N2480,N2800} memory model (guaranteeing sequential
consistency for data-race free programs). Being conservative, delegate
operations block the calling worker until they have completed globally,
ensuring they can be used for synchronization. As such, delegate operations
within a task are guaranteed to be globally visible in program order and all
tasks observe the same global order. Because it is not immediately obvious
that this distributed version of flat combining preserves sequential
consistency, we will now argue why it does so.

@@

The guarantees of combining set and map operations is more nuanced.
Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally (i.e., we assume data-race freedom).
Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
Note that, if an insert finds that its key is already in the combiner, it does not need to send its own message. However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened, effectively reordering operations.
In the same way, lookups can piggy-back on other lookups of the same key.
It is tempting to allow a lookup to be satisfied locally by inspecting the keys to be inserted.
However, this allows the local order in which keys were inserted to be observed.
If this is allowed, then that same order must be respected globally, forcing each batch to ensure atomicity from all other cores' batches. This is prohibitively expensive, so a cheaper solution is chosen: lookups must get their values from the global data structure, but can be performed in parallel in the same way as inserts.

@@

These requirements only guarantee serializability for operations on a data structure in isolation. To guarantee sequential consistency with respect to accesses to all data structures, data race freedom  must be guaranteed externally, just as in any multithreaded shared-memory program.
